---
title: "Pitching Dashboard"
author: "Sean Henderson"
date: '2022-10-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r loading relevant packages and data}
library(tidyverse)
library(readxl)
library(ggplot2)
library(simputation)
pitch_data <- read_excel("2023 Analytics Internship Problem Dataset.xlsx")
```

```{r filtering dataset to use}
pitch_data <- pitch_data|> filter(PITCH_TYPE_KEY != "UN", PITCH_TYPE_KEY != "NULL" , PITCH_NUMBER != "NULL")
pitch_data <- pitch_data|> 
  mutate(woba_value = case_when(EVENT_RESULT_KEY == "single" ~ .9,
                                EVENT_RESULT_KEY == "double" ~ 1.25,
                                EVENT_RESULT_KEY == "triple" ~ 1.6,
                                EVENT_RESULT_KEY == "home_run" ~ 2,
                                (PITCH_RESULT_KEY == "InPlay" & pitch_data$O > 0) ~ 0)) |> 
  mutate(GRAVITY = 36)

pitch_data$RELEASE_HEIGHT <- pitch_data$RELEASE_HEIGHT * 12
pitch_data$PLATE_Z <- pitch_data$PLATE_Z * 12
pitch_data$PLATE_X <- pitch_data$PLATE_X * 12


two_Pitchers <- pitch_data|>
  select(PITCHER_KEY,PITCH_TYPE_KEY,PLATE_X,PLATE_Z,PITCH_RESULT_KEY,RELEASE_SPEED,
         HORIZONTAL_BREAK,INDUCED_VERTICAL_BREAK,RELEASE_HEIGHT,woba_value,GRAVITY) |>
  filter(PITCHER_KEY == "A"  | PITCHER_KEY == "B" , PITCH_TYPE_KEY != "SI")

two_Pitchers$INDUCED_VERTICAL_BREAK <- as.numeric(as.character(two_Pitchers$INDUCED_VERTICAL_BREAK))
da1 <- impute_lm(two_Pitchers, INDUCED_VERTICAL_BREAK ~ RELEASE_HEIGHT-GRAVITY-PLATE_Z)
two_Pitchers <- da1
```
## Pitch Breakdown

This plot shows the breakdown of types of pitches thrown by each pitcher. Both pitchers heavily rely on their fastball, with a mix of a secondary breaking ball. Their 3rd-5th pitches are not thrown frequent enough to be relied upon to get them out of a jam, but still available in their arsenal. 

```{r pitch breakdown, echo=FALSE}
two_PitcherAmounts <- two_Pitchers |> 
    group_by(PITCHER_KEY,PITCH_TYPE_KEY) |> 
  summarize(AmountOfPitches = n())
  
ggplot(two_PitcherAmounts, aes(x = PITCHER_KEY, y = AmountOfPitches, 
                          fill = two_PitcherAmounts$PITCH_TYPE_KEY)) + 
geom_bar(stat = "identity", position = "dodge")
```

## Velocity Plot

This plot shows the comparison of velocities between pitchers for each type of pitch. For all pitches but the curveball, pitcher A has a clear velocity advantage. In addition, pitcher A has better velocity control over his pitches compared to pitcher B, especially when looking at the curveball and four-seamer. 



```{r velocity, echo=FALSE}
two_Pitchers  |> 
ggplot(two_Pitchers,mapping = aes(PITCHER_KEY,RELEASE_SPEED)) + geom_jitter() + 
  facet_wrap(~ two_Pitchers$PITCH_TYPE_KEY)
```
## Strikezone plot

This plot shows the strikezone locaton for each type of pitch. Pitcher B has less horizontal control with his changeup and his curveball compared to pitcher A. Pitcher A utilizes sliders outside the strike zone more often than pitcher B.  


```{r strikezone, echo=FALSE}
ggplot(two_Pitchers,aes(PLATE_X,PLATE_Z)) + geom_point() + facet_grid(two_Pitchers$PITCH_TYPE_KEY ~ PITCHER_KEY) + 
  geom_rect(mapping = aes(ymax = 40.68,ymin = 18.6, xmax = -12, xmin = 12), alpha = 0, size = 1.2, colour = "red")
```

## movement plot

This plot compares the vertical and horizontal movement for each type of pitch. The fastball movement was very similar for both pitchers. Pitcher A was able to generate more vertical break on his curveball compared to pitcher B. One thing that caught my eye was the lack of break from pitcher A's slider, I was expecting more of a general horizontal break. 

```{r movement, echo=FALSE}
ggplot(two_Pitchers,aes(HORIZONTAL_BREAK,INDUCED_VERTICAL_BREAK,col = PITCHER_KEY))+
  geom_point() + facet_wrap(~ two_Pitchers$PITCH_TYPE_KEY) 
```
## Woba Plot

This plot shows all the possible outcomes for balls that were in play. Pitcher B's changeup to the right of the strike zone was not favorable for him, having 4 of 5 pitches being singles. Pitcher A's fastball in the bottom right of the plate wasn't a desirable location for him, with frequent base hits off pitches there. Pitcher A should throw his slider more frequently in the bottom right of the plate, having an approximate woba of .042 in that location.  

```{r woba, echo = FALSE}

two_Pitchers_inPlay <- two_Pitchers
two_Pitchers_inPlay <- two_Pitchers_inPlay |> 
  filter(PITCH_RESULT_KEY == "InPlay" ,PITCH_TYPE_KEY != "FF")

ggplot(two_Pitchers_inPlay, aes(PLATE_X,PLATE_Z)) + geom_point(aes(color = as.character(woba_value))) + 
facet_grid(two_Pitchers_inPlay$PITCH_TYPE_KEY ~ two_Pitchers_inPlay$PITCHER_KEY) + 
geom_rect(mapping = aes(ymax = 40.68, ymin = 18.6, xmax = -12, xmin = 12), alpha = 0, size = 1.2, colour = "black")
```